---
title: "Report on Income Trends"
author: "Justin Lesniak"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(include = FALSE)


knitr::opts_chunk$set(fig.width = 10, fig.height = 5)
```


```{r load packages and data}
library(tidyverse)
library(gtsummary)
library(survey)
library(purrr)
library(here)

source("E:/Research Projects/Frank Chapter 3/Code/R/check_and_fix_survey_ids.R")

load(here(file = "./Processed Data/Individual_quarter_ENCFT_clean.rda" ))

```


```{r prepare data and set up survey object}


#function to check if PSU and STRATA ID repeat across years. If yes returns information on overlap and new data with unique PSU/STRATA ID
check <- check_and_fix_survey_ids(ENCFT_quarterly_individual, psu_var = "UPM", strata_var = "ESTRATO", time_var = "year_quarter")

survey_data_fixed_id <- check$data



#define survey object
survey_design_adj <- svydesign(id = ~psu_unique , weights = ~FACTOR_EXPANSION, strata = ~strata_unique, data = survey_data_fixed_id,
                               nest = TRUE)


#define breaks
cat <- levels(survey_data_fixed_id$year_quarter)
selected_breaks <- cat[seq(1, length(cat), by = 2)]
  

```

# Sources of Income
When looking at wage income most workers get the bulk of their pay from their primary income. Secondary income makes up a small share of income and is stable while primary income has been growing over time.

```{r Sources of Income chart Code}

###### Part 1. Assessing Sources of Income in the Full Labor Force ######


## Plot Share of Labor Force Receiving Income ##

plot_data <-svyby(~has_primary_income + has_secondary_income + has_income, ~year_quarter, design = subset(survey_design_adj, PEA == 1) , svymean)


p_inc_shares <- plot_data %>%
  select(-starts_with("se")) %>%
  pivot_longer( cols = -c("year_quarter"),
                names_to = "income_type",
                values_to = "shares"
  ) %>%
  ggplot(aes(x = year_quarter, y = shares, color = income_type, group = income_type)) +
  geom_line() +
  labs(
    x = "Period",
    y = "Share",
    title = "Share of Labor Force Receiving Income",
    subtitle = "Share of Economically Active Population",
    color = "Income Type") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Average Income of Labor Force Participants ##

plot_data <-svyby(~total_income + primary_income + secondary_income, ~year_quarter, design = subset(survey_design_adj, PEA == 1) , svymean)


p_inc_amounts <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -c("year_quarter"),
                names_to = "income_type",
                values_to = "amount"
  ) %>%
  ggplot(aes(x = year_quarter, y = amount, color = income_type, group = income_type)) +
  geom_line() +
  labs(
    x = "Period",
    y = "Amount",
    title = "Average Income of Labor Force by Type",
    subtitle = "Domincan Pesos",
    color = "Income Type") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))


## Plot Income Shares##

p_income_share <- plot_data %>%
  select(-starts_with("se.") ) %>%
  mutate(total_income = primary_income + secondary_income,
         share_primary = primary_income / total_income,
         share_secondary = secondary_income / total_income ) %>%
  pivot_longer(cols = c("share_primary", "share_secondary") ,
               names_to = "Income_Source",
               values_to = "share") %>%
  ggplot( aes(x = year_quarter, y = share, fill = Income_Source)) +
  geom_bar( position = "stack", stat = "identity")+
  labs(
    x = "Period",
    y = "share",
    title = "Share of Income From Different Sources",
    subtitle = "Percent of Average Monthly Income in a Quarter",
    fill = "Income Source") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))

```

```{r Source of Income Trend Plots, include = TRUE}
plot(p_inc_shares)
plot(p_inc_amounts)
plot(p_income_share)
```


# Income Trends Across the Labor Force

Looking at primary income across groups reveals several interesting trends. Formal workers have much higher income than informal workers and their income is growing while informal salaries are roughly constant.

In terms of worker characteristics, Women in the labor force seem to make higher wages on average than men and there is a significant premium for university employees.



```{r Code for Income Trends of Groups}

###### Part 2 Assessing Income by Worker Traits ######



## Plot Avg Income in Formal and Informal Work ##
plot_data <-svyby(~primary_income, ~year_quarter + Employment, design = subset(survey_design_adj, PEA == 1) , svymean) %>%
  filter(!(Employment == "No Work"))

p_avg_income_by_emp_type <- plot_data %>%
  select(-starts_with(".se")) %>%
  ggplot(aes(x = year_quarter, y = primary_income, color = Employment, group = Employment)) + 
  geom_line() +
  labs(
    x = "Period",
    y = "Amount",
    title = "Average Income of Labor Force by Type",
    subtitle = "Domincan Pesos",
    color = "Employment_Type") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))


## Plot Avg Income by Sector##
plot_data <-svyby(~primary_income, ~year_quarter + Employment_Sector, design = subset(survey_design_adj, PEA == 1) , svymean)


p_avg_income_by_sector <- plot_data %>%
  select(-starts_with(".se")) %>%
  ggplot(aes(x = year_quarter, y = primary_income, color = Employment_Sector, group = Employment_Sector)) + 
  geom_line() +
  labs(
    x = "Period",
    y = "Amount",
    title = "Average Income of Labor Force by Economic Sector",
    subtitle = "Domincan Pesos",
    color = "Employment Sector") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))


## Plot Avg Income by Type of Role##
plot_data <-svyby(~primary_income, ~year_quarter + Employment_Type, design = subset(survey_design_adj, PEA == 1) , svymean)

p_avg_income_by_job_type <- plot_data %>%
  select(-starts_with(".se")) %>%
  ggplot(aes(x = year_quarter, y = primary_income, color = Employment_Type, group = Employment_Type)) + 
  geom_line() +
  labs(
    x = "Period",
    y = "Amount",
    title = "Average Income of Labor Force by Job Type",
    subtitle = "Domincan Pesos",
    color = "Employment Type") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Avg Income by Gender ##
plot_data <-svyby(~primary_income, ~year_quarter + Sex, design = subset(survey_design_adj, PEA == 1) , svymean)


p_avg_income_by_sex <- plot_data %>%
  select(-starts_with(".se")) %>%
  ggplot(aes(x = year_quarter, y = primary_income, color = Sex, group = Sex)) + 
  geom_line() +
  labs(
    x = "Period",
    y = "Amount",
    title = "Average Income of Labor Force by Sex",
    subtitle = "Domincan Pesos",
    color = "Sex") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Avg Income by Education ##
plot_data <-svyby(~primary_income, ~year_quarter + education, design = subset(survey_design_adj, PEA == 1) , svymean)


p_avg_income_by_education <- plot_data %>%
  select(-starts_with(".se")) %>%
  ggplot(aes(x = year_quarter, y = primary_income, color = education, group = education)) + 
  geom_line() +
  labs(
    x = "Period",
    y = "Amount",
    title = "Average Income of Labor Force by Education Level",
    subtitle = "Domincan Pesos",
    color = "Education") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))




```


```{r Income Trend in Overall Labor Force Plots, include = TRUE}
plot(p_avg_income_by_emp_type)
plot(p_avg_income_by_job_type)
plot(p_avg_income_by_sector)


plot(p_avg_income_by_sex)
plot(p_avg_income_by_education)
```




