---
title: "Data on the Labor Force in Dominican Republic"
author: "Justin Lesniak"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(include = FALSE)


knitr::opts_chunk$set(fig.width = 10, fig.height = 5)

```



```{r load packages and data}
library(tidyverse)
library(gtsummary)
library(survey)
library(purrr)
library(here)

source("E:/Research Projects/Frank Chapter 3/Code/R/check_and_fix_survey_ids.R")

load(here(file = "./Processed Data/Individual_quarter_ENCFT_clean.rda" ))


```



```{r prepare data and set up survey object}


#function to check if PSU and STRATA ID repeat across years. If yes returns information on overlap and new data with unique PSU/STRATA ID
check <- check_and_fix_survey_ids(ENCFT_quarterly_individual, psu_var = "UPM", strata_var = "ESTRATO", time_var = "year_quarter")

survey_data_fixed_id <- check$data



#define survey object
survey_design_adj <- svydesign(id = ~psu_unique , weights = ~FACTOR_EXPANSION, strata = ~strata_unique, data = survey_data_fixed_id,
                               nest = TRUE)

#define breaks
cat <- levels(survey_data_fixed_id$year_quarter)
selected_breaks <- cat[seq(1, length(cat), by = 2)]
```
# General Trends in the Labor Market

Overall, we can see at a macro level labor force participation had been trending up pre-Covid, experienced a drop and seems to have recovered. 

Leading up to Covid there was a slight rise in formal work and a decline in informality, after Covid this seems to have reversed. Clear majorities of the workforce are either self-employed or employed in the private sector and this is fairly stable over time. Largest sectors of employment are commerce and the rest of the service sector.

There has been a slight increase in the share of the labor force that is female. Education levels have been increasing over time with more people completing secondary education but not necessarily university level education.




```{r General Labor Market code }



## Plot Participation Rate ##
plot_data <- svyby(~PEA + INACTIVO, ~year_quarter, subset(survey_design_adj, age >=15 & age<=64), svymean, na.rm = TRUE) %>%
  rename(active_share = PEA,
         inactive_share = INACTIVO)

p_paricipation_rate <- plot_data %>%
  select(year_quarter, active_share) %>%
  ggplot( aes(x = year_quarter, y = active_share, group = 1)) +
  geom_line() +
  labs(
    x = "Period",
    y = "Share",
    title = "Economically Active Population",
    subtitle = "Share of Total Population 15-64") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))





## Plot Unemployment Rate ##
plot_data <- svyby(~OCUPADO + DESOCUPADO + SUBOCUPADO, ~year_quarter, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE) %>%
  rename(emp_share = OCUPADO,
         unemp_share = DESOCUPADO,
         underemp_share = SUBOCUPADO
         )



p_unemployment_rate <- plot_data %>%
  select(year_quarter, unemp_share) %>%
  ggplot( aes(x = year_quarter, y = unemp_share, group = 1)) +
  geom_line() +
  labs(
    x = "Period",
    y = "Share",
    title = "Unemployed Population",
    subtitle = "Share of Laborforce") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12))




## Plot share formal/informal workers ##
plot_data <- svyby(~Employment,  ~year_quarter, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment")


p_formal_informal_emp <- plot_data %>%
  select(year_quarter, Formal, Informal) %>%
  pivot_longer( cols = c("Formal", "Informal"),
                names_to = "employment_type",
                values_to = "shares"
  ) %>%
  ggplot(aes(x = year_quarter, y = shares, color = employment_type, group = employment_type)) +
  geom_line() +
  labs(
    x = "Period",
    y = "Share",
    title = "Informal and Formal population over time",
    subtitle = "Share of Employed Population",
    color = "Employment Type") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Share of Workers in Each Category ##
plot_data <- svyby(~Employment_Type,  ~year_quarter, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment_Type")


#plot category shares of labor force
p_category_emp <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -year_quarter,
                names_to = "employment_category",
                values_to = "shares"
  ) %>%
  mutate(employment_category = fct_reorder(employment_category, shares)) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = employment_category)) +
  geom_bar(position = "fill", stat = "identity") +
  labs(
    x = "Period",
    y = "Share",
    title = "Category of Workers",
    subtitle = "Share of Labor Force",
    fill = "Worker Category") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12))


## Plot Share of Workers in Each Sector of Employment ##
plot_data <- svyby(~Employment_Sector,  ~year_quarter, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment_Sector")

p_sector_emp <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -year_quarter,
                names_to = "employment_sector",
                values_to = "shares"
  ) %>%
  mutate(employment_sector = fct_reorder(employment_sector, shares)) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = employment_sector)) +
  geom_bar(position = "stack", stat = "identity") +
  labs(
    x = "Period",
    y = "Share",
    title = "Employment Share by Sector",
    subtitle = "Share of Labor Force",
    fill = "Employment Sector") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12))





## Plot Workers by Gender ##
plot_data <- svyby(~Sex, ~year_quarter, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Sex")

p_labor_force_by_gender <- plot_data %>%
  select(year_quarter, Male, Female) %>%
  pivot_longer( cols = c("Male", "Female"),
                names_to = "Sex",
                values_to =  "Share"
  ) %>%
  ggplot( aes( x = year_quarter, y = Share, color = Sex, group = Sex)) +
  geom_line() +
  labs(
    x = "Period",
    y = "Share",
    title = "Breakdown of Labor Force by Sex",
    subtitle = "Share of Labor Force",
    color = "Sex") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Workers by Education ##
plot_data <- svyby(~education,  ~year_quarter, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "education")


#plot education shares of labor force
p_education_emp <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -year_quarter,
                names_to = "employment_sector",
                values_to = "shares"
  ) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = fct_rev(employment_sector))) +
  geom_bar(position = "fill", stat = "identity") +
  labs(
    x = "Period",
    y = "Share",
    title = "Education Levels in the Labor Force",
    subtitle = "Share of Labor Force",
    fill = "Education Level") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12))


## Plot Average Incomes ##

#Income among employed persons
plot_data <- svyby(~primary_income + secondary_income, ~year_quarter, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE) %>%
  rename(primary_salary = primary_income,
         secondary_salary = secondary_income)


p_income_amt <- plot_data %>%
  select(-starts_with("se.") ) %>%
  pivot_longer(cols = -c(year_quarter), 
               names_to = "Income_Source",
               values_to = "Income") %>%
  ggplot( aes(x = year_quarter, y = Income, color = Income_Source, group = Income_Source)) +
  geom_line() +
  labs(
    x = "Period",
    y = "Amount",
    title = "Average Income for Employed Persons",
    subtitle = "Amount in Dominican Pesos, Average Monthly Income in a Quarter",
    group = "Income Source") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))


## Plot Income Shares##

p_income_share <- plot_data %>%
  select(-starts_with("se.") ) %>%
  mutate(total_income = primary_salary + secondary_salary,
         share_primary = primary_salary / total_income,
         share_secondary = secondary_salary / total_income ) %>%
  pivot_longer(cols = c("share_primary", "share_secondary") ,
               names_to = "Income_Source",
               values_to = "share") %>%
  ggplot( aes(x = year_quarter, y = share, fill = Income_Source)) +
  geom_bar( position = "stack", stat = "identity")+
  labs(
    x = "Period",
    y = "share",
    title = "Share of Income From Different Sources",
    subtitle = "Percent of Average Monthly Income in a Quarter",
    fill = "Income Source") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



```

## Macro Trends

```{r labor market macro trend plots, include = TRUE}
plot(p_paricipation_rate)
plot(p_unemployment_rate)
```

## Trends in Employment Type

```{r labor market Type of Work plots, include = TRUE}
plot(p_formal_informal_emp)
plot(p_category_emp)
plot(p_sector_emp)

```
## Trends in Worker Characteristics

```{r labor market Worker Traits, include = TRUE}
plot(p_labor_force_by_gender)
plot(p_education_emp)
plot(p_income_amt)
plot(p_income_share)
```




# Trends in Formal vs Informal Sector Workers
Comparing formal and informal sector we can see most workers with formal employment are classified as either private or public sector employees. In the informal sector there is are a significant number of private sector employees but the largest group is self-employed which makes up a negligible amount of formal employment. Looking at economic sectors, we see commerce and rest of services is a bit more dominant in the informal sector while the formal sector has a more even mix of sectors.

According to the survey men make up the majority of both types of workers but have even greater representation in the informal sector. Looking at education levels, the formal sector workers have secondary or higher education for the most part while in the informal sector the majority have primary education although this share is falling and being replaced by workers with secondary education over time.



```{r Formal vs Informal Sector Charts}

## Plot Type of Worker in Formal vs Informal Sectors ##


plot_data <- svyby(~Employment_Type,  ~year_quarter + Employment, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment_Type") %>%
  filter(!(Employment == "No Work"))

p_form_inform_by_cat <- plot_data %>%
  filter(!(Employment == "No Work")) %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -c(year_quarter, Employment),
                names_to = "employment_category",
                values_to = "shares"
  ) %>%
  mutate(employment_category = fct_reorder(employment_category, shares)) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = employment_category)) +
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(~Employment) +
  labs(
    x = "Period",
    y = "Share",
    title = "Type of Workers in the Formal and Informal Sector",
    subtitle = "Share of Group",
    fill = "Worker Category") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Major Sectors of Employment in Formal vs Informal Sectors ##

plot_data <- svyby(~Employment_Sector,  ~year_quarter + Employment, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment_Sector") %>%
  filter(!(Employment == "No Work"))

p_form_inform_by_sector <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -c(year_quarter, Employment),
                names_to = "employment_sector",
                values_to = "shares"
  ) %>%
  mutate(employment_sector = fct_reorder(employment_sector, shares)) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = employment_sector)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_wrap(~Employment) +
  labs(
    x = "Period",
    y = "Share",
    title = "Sectoral Employment Shares in the Formal and Informal Sector",
    subtitle = "Share of Group",
    fill = "Employment Sector") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Gender Comparison of Formal vs Informal Sectors ##

plot_data <- svyby(~Sex, ~year_quarter + Employment, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Sex") %>%
  filter(!(Employment == "No Work"))



p_form_inform_by_gender <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = c(Male, Female),
                names_to = "Sex",
                values_to =  "share"
  ) %>%
  ggplot( aes( x = year_quarter, y = share, fill = Sex)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_wrap(~Employment) +
  labs(
    x = "Period",
    y = "Share",
    title = "Sex Breakdown in Formal Versus Informal Sector",
    subtitle = "Share of Group",
    fill = "Sex") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Education Comparison of Formal vs Informal Sectors ##

plot_data <- svyby(~education,  ~year_quarter + Employment, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "education") %>%
  filter(!(Employment == "No Work"))


#plot education shares of labor force
p_form_inform_by_educ <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -c(year_quarter, Employment),
                names_to = "eduction",
                values_to = "shares"
  ) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = fct_rev(eduction))) +
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(~Employment) + 
  labs(
    x = "Period",
    y = "Share",
    title = "Education Levels in Formal vs Informal Sector",
    subtitle = "Share of Group",
    fill = "Education Level") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))

## Plot Income of Formal/Informal workers ##


#Income among employed persons
plot_data <- svyby(~primary_income + secondary_income, ~year_quarter + Employment, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE) %>%
  rename(primary_salary = primary_income,
         secondary_salary = secondary_income)


#Income amounts
p_income_amt_by_type <- plot_data %>%
  select(-starts_with("se.") ) %>%
  pivot_longer(cols = -c(year_quarter, Employment), 
               names_to = "Income_Source",
               values_to = "Income") %>%
  ggplot( aes(x = year_quarter, y = Income, color = Income_Source, group = Income_Source)) +
  geom_line() +
  facet_wrap(~Employment) +
  labs(
    x = "Period",
    y = "Amount",
    title = "Income by Employment Type",
    subtitle = "Amount in Dominican Pesos, Average Monthly Income in a Quarter",
    group = "Income Source") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))


## Plot Income Shares of Formal/Informal workers ##


p_income_share_by_type <- plot_data %>%
  select(-starts_with("se.") ) %>%
  mutate(total_income = primary_salary + secondary_salary,
         share_primary = primary_salary / total_income,
         share_secondary = secondary_salary / total_income ) %>%
  pivot_longer(cols = c("share_primary", "share_secondary") ,
               names_to = "Income_Source",
               values_to = "share") %>%
  ggplot( aes(x = year_quarter, y = share, fill = Income_Source)) +
  geom_bar( position = "stack", stat = "identity") +
  facet_wrap(~Employment) +
  labs(
    x = "Period",
    y = "share",
    title = "Share of Income From Different Sources by Employment Type",
    subtitle = "Percent of Average Monthly Income in a Quarter",
    fill = "Income Source") +
  scale_x_discrete(breaks = selected_breaks) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))

```


## Trends in Employment Type Within the Formal vs Informal Sector
```{r Formal vs Informal Employment Types, include = TRUE}
plot(p_form_inform_by_cat)
plot(p_form_inform_by_sector)
```

## Trends in Worker Traits and Outcomes Within the Formal vs Informal Sector
```{r Formal vs Informal Worker Traits, include = TRUE}
plot(p_form_inform_by_gender)
plot(p_form_inform_by_educ)

plot(p_income_amt_by_type)
plot(p_income_share_by_type)
```


## Labor Force Trends by Sex
Looking at the labor force in terms of sex we see that men seem to be more successful than women. The male participation rate is about 15% higher than the female participation rate and the unemployment rate is almost half for men compared to women in most years.

In terms of type of employment, we can see men have slightly more informal work and less formal work compared to women. We can also see that men have a higher share of people self-employed while women have more employment that is unclassified in the survey. Looking at the sector data, the service sectors are a bit more dominate for women than men.

A higher share of women than men have university education and yet their salary is significantly lower than men on average. There seems to be a similar trend in salary increases for both groups.


```{r Code for Labor Market Gender Graphs}

## Plot Male/Female Participation Rate ###
plot_data <- svyby(~PEA, ~Sex + year_quarter,subset(survey_design_adj, age >=15 & age<=64), svymean, na.rm = TRUE)


p_participation_by_sex <- plot_data %>%
  select(year_quarter, PEA, Sex) %>%
  ggplot( aes( x = year_quarter, y = PEA, fill = Sex)) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(
    x = "Period",
    y = "Share",
    title = "Participation Rate by Sex",
    subtitle = "Share of Working Age Population in Each Sex",
    color = "Sex") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12))





## Plot Male and Female Unemployment Rate ##
plot_data <- svyby(~DESOCUPADO, ~Sex + year_quarter, subset(survey_design_adj, PEA == 1) , svymean, na.rm = TRUE)


p_unemp_by_sex <- plot_data %>%
  select(year_quarter, DESOCUPADO, Sex) %>%
  ggplot( aes( x = year_quarter, y = DESOCUPADO, fill = Sex)) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(
    x = "Period",
    y = "Share",
    title = "Unemployment Rate by Sex",
    subtitle = "Share of Labor Force in Each Sex",
    color = "Sex") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12))








## Plot Male and Female Formal vs Informal Employment ##

plot_data <- svyby(~Employment, ~Sex + year_quarter, subset(survey_design_adj, PEA == 1) , svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment") %>% 
  select(-`No Work`)

p_typework_by_sex <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer(cols = -c(year_quarter, Sex),
                names_to = "work_type",
                values_to = "share") %>%
  ggplot( aes( x = year_quarter, y = share, fill = work_type)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_wrap(~Sex) +
  labs(
    x = "Period",
    y = "Share",
    title = "Employment Type by Sex",
    subtitle = "Share of Employment Type",
    fill = "Work Type") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Type of Workers by Gender ##
plot_data <- svyby(~Employment_Type,  ~year_quarter + Sex, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment_Type")


p_cat_by_sex <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -c(year_quarter, Sex),
                names_to = "employment_category",
                values_to = "shares"
  ) %>%
  mutate(employment_category = fct_reorder(employment_category, shares)) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = employment_category)) +
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(~Sex) +
  labs(
    x = "Period",
    y = "Share",
    title = "Job Category by Sex",
    subtitle = "Share of of Group",
    fill = "Worker Category") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Sectoral Employment by Gender ##
plot_data <- svyby(~Employment_Sector,  ~year_quarter + Sex, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment_Sector")

p_sector_by_sex <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -c(year_quarter, Sex),
                names_to = "employment_sector",
                values_to = "shares"
  ) %>%
  mutate(employment_sector = fct_reorder(employment_sector, shares)) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = employment_sector)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_wrap(~Sex) +
  labs(
    x = "Period",
    y = "Share",
    title = "Sectoral Employment Shares by Sex",
    subtitle = "Share of Group",
    fill = "Employment Sector") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



## Plot Education by Gender ##

plot_data <- svyby(~education,  ~year_quarter + Sex, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "education")


#plot education shares of labor force
p_educ_by_sex <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -c(year_quarter, Sex),
                names_to = "eduction",
                values_to = "shares"
  ) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = fct_rev(eduction))) +
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(~Sex) + 
  labs(
    x = "Period",
    y = "Share",
    title = "Education Levels by Sex",
    subtitle = "Share of Group",
    fill = "Education Level") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))

## Plot Income Levels By Sex ##


plot_data <- svyby(~primary_income + secondary_income, ~year_quarter + Sex, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE) %>%
  rename(primary_salary = primary_income,
         secondary_salary = secondary_income)


p_income_amt_by_sex <- plot_data %>%
  select(-starts_with("se.") ) %>%
  pivot_longer(cols = -c(year_quarter, Sex), 
               names_to = "Income_Source",
               values_to = "Income") %>%
  ggplot( aes(x = year_quarter, y = Income, color = Income_Source, group = Income_Source)) +
  geom_line() +
  facet_wrap(~Sex) +
  labs(
    x = "Period",
    y = "Amount",
    title = "Income by Sex",
    subtitle = "Amount in Dominican Pesos, Average Monthly Income in a Quarter",
    group = "Income Source") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))


## Plot Income Shares By Sex ##

p_income_share_by_sex <- plot_data %>%
  select(-starts_with("se.") ) %>%
  mutate(total_income = primary_salary + secondary_salary,
         share_primary = primary_salary / total_income,
         share_secondary = secondary_salary / total_income ) %>%
  pivot_longer(cols = c("share_primary", "share_secondary") ,
               names_to = "Income_Source",
               values_to = "share") %>%
  ggplot( aes(x = year_quarter, y = share, fill = Income_Source)) +
  geom_bar( position = "stack", stat = "identity") +
  facet_wrap(~Sex) +
  labs(
    x = "Period",
    y = "share",
    title = "Share of Income From Different Sources by Sex",
    subtitle = "Percent of Average Monthly Income in a Quarter",
    fill = "Income Source") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))


```


## Macro Trends in the Labor Market by Sex
```{r Labor Market Trends by Sex, include = TRUE}
plot(p_participation_by_sex)
plot(p_unemp_by_sex)
```

## Trends in Employment type by Sex
```{r Labor Market Job Type by Sex, include = TRUE}
plot(p_typework_by_sex)
plot(p_cat_by_sex)
plot(p_sector_by_sex)
```

## Trends in Worker Traits and Outcomes by Sex
```{r Labor Market Worker Outcomes by Sex, include = TRUE}
plot(p_educ_by_sex)
plot(p_income_amt_by_sex)
plot(p_income_share_by_sex)
```

## Labor Force Trends by Education Level

Looking at the labor force in terms of education it appears participation rates go up substantially with education as do unemployment rates.

More educated groups have a higher proportion of members working in the formal sector as opposed to the informal sector. In terms of job category, we see that the share of people working in "unclassified" jobs according to the survey drops substantially with education level. We also see that there is a greater diversity of sectoral work in the higher education groups.

In terms of income there appears to be a significant premium for secondary school graduates and an even greater increase for university graduates relative to those with only primary education.

```{r Education Charts Code}

## Participation Rate by education group ###
plot_data <- svyby(~PEA, ~education + year_quarter, subset(survey_design_adj, age >=15 & age<=64), svymean, na.rm = TRUE)


p_participation_by_edu <- plot_data %>%
  select(year_quarter, PEA, education) %>%
  ggplot( aes( x = year_quarter, y = PEA, color = education, group = education)) +
  geom_line() +
  labs(
    x = "Period",
    y = "Share",
    title = "Participation Rate by Education",
    subtitle = "Share of Working Age Population in Each Group that is Economically Active",
    color = "Education Level") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12))





##Unemployment Rate by education group ##
plot_data <- svyby(~DESOCUPADO, ~year_quarter + education , subset(survey_design_adj, PEA == 1) , svymean, na.rm = TRUE)


p_unemp_by_edu <- plot_data %>%
  select(year_quarter, DESOCUPADO, education) %>%
  ggplot( aes( x = year_quarter, y = DESOCUPADO, color = education, group = education)) +
  geom_line() +
  labs(
    x = "Period",
    y = "Share",
    title = "Unemployment Rate by Education Level",
    subtitle = "Share of Population Unemployed",
    color = "Education Level") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 12))





####Job Trends ####


##share formal/informal workers by education
plot_data <- svyby(~Employment, ~education + year_quarter, subset(survey_design_adj, PEA == 1) , svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment")


p_typework_by_edu <- plot_data %>%
  select(-`No Work`) %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -c(year_quarter, education),
                names_to = "work_type",
                values_to = "share") %>%
  ggplot( aes( x = year_quarter, y = share, fill = work_type)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_wrap(~education) +
  labs(
    x = "Period",
    y = "Share",
    title = "Employment Type by Education",
    subtitle = "Share of Employment Type",
    fill = "Work Type") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



##Category of Workers ##
plot_data <- svyby(~Employment_Type,  ~year_quarter + education, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment_Type")


#plot category shares of labor force
p_cat_by_edu <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -c(year_quarter, education),
                names_to = "employment_category",
                values_to = "shares"
  ) %>%
  mutate(employment_category = fct_reorder(employment_category, shares)) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = employment_category)) +
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(~education) +
  labs(
    x = "Period",
    y = "Share",
    title = "Job Category by Education",
    subtitle = "Share of of Group",
    fill = "Worker Category") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



##Major Sectors of Employment ##
plot_data <- svyby(~Employment_Sector,  ~year_quarter + education, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE)
plot_data <- clean_svy_names(plot_data, "Employment_Sector")

p_sector_by_edu <- plot_data %>%
  select(-starts_with("se.")) %>%
  pivot_longer( cols = -c(year_quarter, education),
                names_to = "employment_sector",
                values_to = "shares"
  ) %>%
  mutate(employment_sector = fct_reorder(employment_sector, shares)) %>%
  ggplot(aes(x = year_quarter, y = shares, fill = employment_sector)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_wrap(~education) +
  labs(
    x = "Period",
    y = "Share",
    title = "Sectoral Employment Shares by Education",
    subtitle = "Share of Group",
    fill = "Employment Sector") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))






## Plot Income Levels By education ##

#Income among employed persons
plot_data <- svyby(~primary_income + secondary_income, ~year_quarter + education, subset(survey_design_adj, PEA == 1), svymean, na.rm = TRUE) %>%
  rename(primary_salary = primary_income,
         secondary_salary = secondary_income)


#Income amounts
p_income_amt_by_edu <- plot_data %>%
  select(-starts_with("se.") ) %>%
  pivot_longer(cols = -c(year_quarter, education), 
               names_to = "Income_Source",
               values_to = "Income") %>%
  ggplot( aes(x = year_quarter, y = Income, color = Income_Source, group = Income_Source)) +
  geom_line() +
  facet_wrap(~education) +
  labs(
    x = "Period",
    y = "Amount",
    title = "Income by Education",
    subtitle = "Amount in Dominican Pesos, Average Monthly Income in a Quarter",
    group = "Income Source") +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))


## Plot Income Shares By education ##

p_income_share_by_edu <- plot_data %>%
  select(-starts_with("se.") ) %>%
  mutate(total_income = primary_salary + secondary_salary,
         share_primary = primary_salary / total_income,
         share_secondary = secondary_salary / total_income ) %>%
  pivot_longer(cols = c("share_primary", "share_secondary") ,
               names_to = "Income_Source",
               values_to = "share") %>%
  ggplot( aes(x = year_quarter, y = share, fill = Income_Source)) +
  geom_bar( position = "stack", stat = "identity") +
  facet_wrap(~education) +
  labs(
    x = "Period",
    y = "share",
    title = "Share of Income From Different Sources by Education",
    subtitle = "Percent of Average Monthly Income in a Quarter",
    fill = "Income Source") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(breaks = selected_breaks) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        strip.text = element_text(size = 12))



```

## Macro Trends in the Labor Force by Education Level
```{r Labor Market Trends by Education, include = TRUE}
plot(p_participation_by_edu)
plot(p_unemp_by_edu)
```

## Trends in Employment type by Education Level
```{r Labor Market Job Type by Education, include = TRUE}
plot(p_typework_by_edu)
plot(p_cat_by_edu)
plot(p_sector_by_edu)
```

## Trends in Worker Traits and Outcomes by Education Level
```{r Labor Market Worker Outcomes by Education, include = TRUE}
plot(p_income_amt_by_edu)
plot(p_income_share_by_edu)
```

